%! Author = borisdeletic
%! Date = 02/04/2023

% Preamble
\documentclass[11pt]{article}

% Packages
\usepackage{amsmath}
\usepackage{biblatex}

\addbibresource{../bib/references.bib}
% Document
\begin{document}

    \section{Introduction}
    \subsection{Motivations (LFT)}
    Lattice field theory can be interpreted as a Bayesian sampling problem.
    Often very difficult due to multimodal likelihoods with deep topological traps.
    Topological freezing is a significant challenge which Nested Sampling helps to overcome.

    \subsection{Hamiltonian Monte Carlo (classical for LFT)}
    Describe classical HMC algorithm. \autocite{betancourt2018conceptual}

    \subsection{Bayesian Inference}
    Introduction to bayesian inference.

    \subsection{Nested Sampling}
    Sampling with Hard Likelihood constraint.

    \section{Constrained HMC for Nested Sampling}
    \subsection{Reflections}
    Compression of posterior as we shrink the isolikelihood contour.
    Cannot pick a static value for epsilon and metric which is effective for sampling
    from large posterior space and also the compressed space.
    This necessitates a dynamic adaption of our HMC parameters (epsilon, metric).

    \section{Parameter Adaption}
    \subsection{Epsilon Dual Averaging}
    \subsection{Metric Scaling using Energy}
    \subsection{Reflection Rates for local minima}

    \section{Results}
    \subsection{Phi4}
    \subsection{Topological Trap}
    \subsection{Performance}


    \appendix
    \section{Code Design}
    \subsection{Interfaces}
    \subsection{Dependency Injection}
    \subsection{Unit Tests}
    \subsection{Autodiff}

\printbibliography
\end{document}